SUPPLEMENTARY:\\A new approach to bias correction in high-throughput sequencing data

### 2011-09-30 seqbias-supplement.tex
final draft

\title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
\author{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}

@@

\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
Sequencing Data}}{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-sampstd.pdf}
\end{center}
\caption{Sensitivity of the parameter control he standard deviation at which
background sequences are sampled.}
\label{fig:poissampstd}
\end{figure}

Figure \ref{fig:poissampstd} show the standard deviation used to draw background
samples varied from 1 to 10,000. Apparent from this plot is that, while the
optimal choice lies somewhere between 10 and 100, the model performs
competitively for any reasonable value.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-comppen.pdf}
\end{center}
\caption{Sensitivity over adjustments to the complexity penalty.}
\label{fig:poiscomppen}
\end{figure}

@@

Figure \ref{fig:caokl}, we plot directly the positional KL divergence, computed
in tho same manner described in the Results section of the paper.  We see that
our method is effective at reducing the bias in this ChIP-Seq data.

@@

It is clear from this analysis that the FRT-Seq method, as implemented to
generate this data, is not without bias, yet there is significant divergence at
only a small number of positions surrounding the read start. When our method is
trained on this data we obtain the suitably sparse model pictured in Figure
\ref{fig:frtgraph}.

@@

Though we do not know if the variability is always minor, assuming this is so,
there remains the risk that the sequence bias would greatly effect the depth to
which a locus is sequenced, and thus the statistical significance of
differential expression tests. This would bias the discovery of differentially
expressed genes, but not the test itself.

\section{KL Divergence}

@@

Suppose $P$ and $Q$ are probability distributions defined on a discrete sample
space. E.g., in the context of this paper, think of them as the probabilities of
DNA $k$-mers for some fixed $k$. The \emph{Kullback-Leibler divergence}, also
known as \emph{relative entropy}, of $Q$ with respect to $P$ is defined as
$$H(Q||P) = \sum_{i} q_i \ln \frac{q_{i}}{p_{i}}$$
where $q_{i}$ ($p_{i}$) is the probability of observing the i$^{\text{th}}$
event according to the distribution $Q$ (resp., $P$), and the summation is taken
over all events in the sample space (e.g., all $k$-mers). In some sense, this is
a measure of the dissimilarity between the distributions: if $p_{i} \approx
q_{i}$ everywhere, their log rations will be near zero and $H$ will be small; as
$q_{i}$ and $p_{i}$ diverge, their log rotions will deviate from zero and $H$
will increase.

For a more quantitative and (perhaps) intuitive interpretation, consider the
following hypothesis-testing scenario. Given $m$ independent samples from either
$P$ (arbitrarily called the ``null'') or from $Q$ (the ``alternative''), how
large should $m$ be to cenfidently choose between these cases? A natural
approach is to use a likelihood ratio test: letting $Y_i$ be the number of times
event $i$ is observed in $m$ trials ($\sum Y_i = m$), the likelihood of the
sequence of observations under model $P$ is $$\prod_{i} p_{i}^{Y_{i}}$$
and similarily for $Q$. So, the logarithm of the likelihood ratios is
$$
\text{LLR} =
\ln \frac{\prod_{i} q_{i} ^ {Y_{i}}}{\prod_{i} p_{i} ^ {Y_{i}}}
= \sum_{i} Y_{i} \ln \frac{q_{i}}{p_{i}}
= m \sum_{i} \frac{Y_{i}}{m} \ln \frac{q_{i}}{p_{i}} $$

If the sample is drawn from the alternative distribution $Q$, then the
expectation of $frac{Y_{i}}{m}$ is $q_i$, so the expected value of the LLR is
$$ m \sum_{i} q_{i} \ln \frac{q_{i}}{p_{i}} = m H(Q || P) $$
That is, the KL divergence is exactly the expected per-sample contribution to
the log-likelihood ratio.
%% when comparing these two hypotheses ??
So, assuming the null hypothesis is false, in order for it to be rejected with
$1000:1$ odds, one should choose $m$ to be inversely proportional to $H(Q ||
P)$:
\begin{align*}
m H(Q||P) \ge \ln 1000 \\
m \ge \frac{\ln 1000}{H(Q || P)}
\end{align*}

As a concrete example, all of the RNA-Seq data sets examined in this paper show
a KL divergence between the 1$^{\text{st}}$ positon of reads versus the
transcriptomic background nucleotide distribution of $0.05$ or greater. Thus,
one can confidently reject the hypothesis that reads are beeing uniformly
sampled across the transcriptome by examining only a few hundred randomly
selected reads (i.e., $\ln 1000 / 0.05 \approx 140$).

@@

Continuing the notation above, suppose $P$ as an unknown distribution with
parameters $p_1, \dots, p_r$, $\sum p_i = 1$ where $r$ is the number of points
in the sample space (e.g. $r = 4^{k}$ in the case of $k$-mers). Given a random
ample $X_1, X_2, \dots, X_r$ of size $n = \sum_{i} X_i$ from $P$, it is well
known that the maximum likelihood estimators for the parameters are $q_i =
\frac{X_i}{n} \approx p_i$. How good an estimate for $P$ is this distribution
$Q$?

$$E[q_i] = E\left[\frac{X_i}{n}\right] = \frac{E[X_i]}{n} = \frac{np_i}{n} =
p_i$$
And the standard deviation of each estimate is proportional to $1/\sqrt{n}$. So
these estimates are increasingly accurate as the sample size increases. A more
quantitative assessment of the accuracy of the estimator is obtained by
evaluating the KL divergence:
$$H(Q||P)
= \sum_{i = 1}^{r} q_i \ln \frac{q_i}{p_i}
= \sum_{i = 1}^{r} q_i \ln \left(1 + \frac{q_i - p_i}{p_i} \right) $$
Using the first two terms of he Taylor series for $\ln (1 + x)$, this is
\begin{align*}
H(Q||P) &\approx \sum_{i = 1}^{r} q_i \left( \frac{q_i - p_i}{p_i} -
\frac{1}{2} \left( \frac{q_i - p_i}{p_i} \right)^2 \right ) \\
&= \sum_{i = 1}^{r} q_i \frac{q_i - p_i}{p_i} -
\frac{q_i}{2 p_i} \frac{(q_i - p_i)^2} {p_i}
\end{align*}
Since $\sum_{i = 1}^{r} q_i = \sum_{i = 1}^{r} pi = 1$,
$\sum_{i = 1}^{r} p_i \frac{q_i - p_i}{p_i} = 0$, so
\begin{align*}
H(Q||P) &\approx \sum_{i=1}^{r} q_i \frac{q_i - p_i}{p_i} - p_i \frac{q_i -
p_i}{p_i} - \frac{q_i}{2 p_i} \frac{(q_i - p_i)^2}{p_i} \\
&= \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}\left(1 - \frac{q_i}{2 p_i}\right) \\
&\approx \frac{1}{2} \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}
\end{align*}
since $q_i \approx p_i$. Multiplying by $n^2 / n^2$ we have,
\begin{align*}
H(Q||P) &\approx \frac{1}{2n} \sum_{i=1}^{r} \frac{n q_i - n p_i)^2}{n p_i} \\
        &= \frac{1}{2n} \sum_{i=1}^{r} \frac{(X_i - E[X_i])^2}{E[X_i]}
\end{align*}

@@

Stochastic simulations of this proves when $P$ is uniform show that the above
formula is a very good fit when $n > r$.

\begin{figure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{fig/entropy.pdf}}
\caption{For each value of $n$ plotted, the circles indicate the value of
$H(Q||P)$, averaged over 100 samples of size $n$, where $P$ is uniform and $Q$
is the MLE estimator for $P$ basen on $n$ random samples drawn from $r$ bins.
The asterisks and the straight line interpolating them are the theoretical
approximation from Equation \ref{eq:entropy}.
}
\label{fig:entropy}
\end{figure}

@@

The training procedure begins by considering the the set of models formed by
including each single nucleotide position from within the training sequences.
For each single nucleotide model, the Bayesian infromation criterion is
evaluated. If in any of these model the BIC score increases over the empty
model, a non-empty model will be trained, otherwise the training procedure will
halt with an empty model.

@@

Now, objective function evaluated is the Bayesian information criterion (BIC)
applied to the conditional log-likelihood (CLL).
$$ \text{BIC} = 2 \cdot \text{CLL} - m \log n $$
where CLL in this single nucleotide case can be written as,
$$ \text{CLL} =
\sum_{k \in \{a, c, g, t\}}
\left(
x_k \log \left( \frac{x_k/n}{(x_k + y_k)/n} \right) +
y_k \log \left( \frac{y_k/n}{(x_k + y_k)/n} \right) \right)$$

@@

In summary, if $D_X, D_Y < \frac{3}{2n} \log n$, then $\text{BIC} \le
\text{BIC}_0$ and the empty model is retained. That is, the probability of an
empty model being trained is bounded below by the probability that the KL
divergence of both sampled is bounded above $\frac{3}{2n} \log n$:
$$ \Pr(D_{X} < \frac{3}{2n} \log n) \cdot 
\Pr(D_{Y} < \frac{3}{2n} \log n) $$

@@

In training we consider mulitple nucleotides. If $h$ nucleotides are considered,
the probability of a non-empty model being trained is less than
$$ 1 - (\Pr(D < \frac{3}{2n} \log n))^{2h} $$

As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $\chi^2$
distribution with 3 degrees of freedom for large $n$, and so we can estimate
this probability for various values of $n$. This is shown in Figure
\ref{fig:nepr}. From this figure we can see, for example that training with more
than 10,000 reads results in an incorrect non-empty model with probability less
than 0.0004.

In the low end, the upper bound on non-empty model probability is much higher.
In our tests, this upper bound is not tight for a small number of reads. We
trained the model on 100 sets of 100 simulated reads, drawn randomly at uniform
from exonic sequences and in only 3 trials was a non-empty model trained, far
less than the upper bound prediction of 23 out of 100. This test was repeated
using sets of 10,000 reads, and in no trial was a non-empty model trained.

From these results, and the results in Supplementary Section 5, we recommend
training our method with at least 10,000 reads.

\begin{figure}[H]
\centerline{
\includegraphics[width=0.8\textwidth]{fig/nepr.pdf}}
\caption{Plotted is the upper bound on the probability of training a non-empty
model when there is no real bias in the experiment when the model is trained
with various number of reads $n$. We fixed the number of nucleotides being
considered as $h = 41$, as this what was used for the evaluation in Section 3.}
\label{fig:nepr}
\end{figure}

>>>>

\title{SUPPLEMENTARY:\\A new approach to bias correction in high-throughput sequencing data}
\author{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}

@@

\dcjtitle{\sc{(SUPPLEMENTARY)}}
{\sc{A new approach to bias correction in high-throughput sequencing data}}
{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-sampstd.pdf}
\end{center}
\caption{Sensitivity of the parameter controling the standard deviation at which
background sequences are sampled, as evaluated with McFadden's $R^2$
goodness-of-fit statistic.}
\label{fig:poissampstd}
\end{figure}

Figure \ref{fig:poissampstd} shows the median McFaddens $R^2$ goodness-of-fit
statastic across 1000 test exons (as described in Section 3.2 of the main paper)
versus the standard deviation used to draw background samples varied from 1 to
10,000. Apparent from this plot is that, while the optimal choice lies somewhere
between 10 and 100, the model performs competitively for any reasonable value.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{fig/pois-comppen.pdf}
\end{center}
\caption{Sensitivity over adjustments to the complexity penalty. The $Y$ axis is
as in Figure \ref{fig:poissampstd}.}
\label{fig:poiscomppen}
\end{figure}

@@

Figure \ref{fig:caokl}, we plot directly the positional KL divergence, computed
in the same manner described in the Results section of the paper.  We see that
our method is effective at reducing the bias in this ChIP-Seq data.

@@

It is clear from this analysis that the FRT-Seq method, as implemented to
generate this data, is not without bias, yet there is significant divergence at
only a small number of positions near the read start, including the position
before it. When our method is trained on this data we obtain the suitably sparse
model pictured in Figure
\ref{fig:frtgraph}.

@@

Though we do not know if the variability is always minor, assuming this is so,
there remains the risk that the sequence bias would greatly effect the depth to
which a locus is sequenced, and thus the statistical significance of
differential expression tests. This would bias the discovery of differentially
expressed genes, but not the test itself.  Additionally, as sequencing bias
appears to be at least partly related to sequence context, alternative splicing
potentially will change the bias at positions somewhat distant from the splice
site itself, complicating any analysis of differential isoform usage.

\section{KL Divergence and Parameter Estimation}

@@

In this section and the next, we present some theoretical analysis quantifying
the accuracy of parameter estimation and the likelihood of falsely learning a
biased model from unbiased data.

Suppose $P$ and $Q$ are probability distributions defined on a discrete sample
space. E.g., in the context of this paper, think of them as the probabilities of
DNA $k$-mers for some fixed $k$. The \emph{Kullback-Leibler divergence}, also
known as \emph{relative entropy}, of $Q$ with respect to $P$ is defined as
$$H(Q||P) = \sum_{i} q_i \ln \frac{q_{i}}{p_{i}}$$
where $q_{i}$ ($p_{i}$) is the probability of observing the i$^{\text{th}}$
event according to the distribution $Q$ (resp., $P$), and the summation is taken
over all events in the sample space (e.g., all $k$-mers). In some sense, this is
a measure of the dissimilarity between the distributions: if $p_{i} \approx
q_{i}$ everywhere, their log ratios will be near zero and $H$ will be small; as
$q_{i}$ and $p_{i}$ diverge, their log ratios will deviate from zero and $H$
will increase.

For a more quantitative and (perhaps) intuitive interpretation, consider the
following hypothesis-testing scenario. Given $m$ independent samples from either
$P$ (arbitrarily called the ``null'') or from $Q$ (the ``alternative''), how
large should $m$ be to confidently choose between these cases? A natural
approach is to use a likelihood ratio test: letting $Y_i$ be the number of times
event $i$ is observed in $m$ trials ($\sum Y_i = m$), the likelihood of the
sequence of observations under model $P$ is $$\prod_{i} p_{i}^{Y_{i}}$$
and similarly for $Q$. So, the logarithm of the likelihood ratios is
$$
\text{LLR} =
\ln \frac{\prod_{i} q_{i} ^ {Y_{i}}}{\prod_{i} p_{i} ^ {Y_{i}}}
= \sum_{i} Y_{i} \ln \frac{q_{i}}{p_{i}}
= m \sum_{i} \frac{Y_{i}}{m} \ln \frac{q_{i}}{p_{i}} $$

If the sample is drawn from the alternative distribution $Q$, then the
expectation of $\frac{Y_{i}}{m}$ is $q_i$, so the expected value of the LLR is
$$ m \sum_{i} q_{i} \ln \frac{q_{i}}{p_{i}} = m H(Q || P) $$
That is, the KL divergence is exactly the expected per-sample contribution to
the log-likelihood ratio.
%% when comparing these two hypotheses ??
So, assuming the null hypothesis is false, in order for it to be rejected with
say, $1000:1$ odds, one should choose $m$ to be inversely proportional to $H(Q ||
P)$:
\begin{align*}
m H(Q||P) \ge \ln 1000 \\
m \ge \frac{\ln 1000}{H(Q || P)}
\end{align*}

As a concrete example, all of the RNA-Seq data sets examined in this paper show
a KL divergence between the 1$^{\text{st}}$ position of reads versus the
transcriptomic background nucleotide distribution of $0.05$ or greater. Thus,
one can confidently reject the hypothesis that reads are beeing uniformly
sampled across the transcriptome by examining only a few hundred randomly
selected reads (i.e., $\ln 1000 / 0.05 \approx 140$), a surprisingly small
number, given the supposedly ``random sampling'' accomplished by RNA-Seq.

@@

Continuing the notation above, suppose $P$ as an unknown distribution with
parameters $p_1, \dots, p_r$, $\sum p_i = 1$ where $r$ is the number of points
in the sample space (e.g. $r = 4^{k}$ in the case of $k$-mers). Given a random
sample $X_1, X_2, \dots, X_r$ of size $n = \sum_{i} X_i$ from $P$, it is well
known that the maximum likelihood estimators for the parameters are $q_i =
\frac{X_i}{n} \approx p_i$. How good an estimate for $P$ is this distribution
$Q$? The estimators are unbiased:
$$E[q_i] = E\left[\frac{X_i}{n}\right] = \frac{E[X_i]}{n} = \frac{np_i}{n} =
p_i$$
and the standard deviation of each estimate is proportional to $1/\sqrt{n}$, so
these estimates are increasingly accurate as the sample size increases. A more
quantitative assessment of the accuracy of the estimator is obtained by
evaluating the KL divergence:
$$H(Q||P)
= \sum_{i = 1}^{r} q_i \ln \frac{q_i}{p_i}
= \sum_{i = 1}^{r} q_i \ln \left(1 + \frac{q_i - p_i}{p_i} \right) $$
Using the first two terms of the Taylor series for $\ln (1 + x)$, this is
\begin{align*}
H(Q||P) &\approx \sum_{i = 1}^{r} q_i \left( \frac{q_i - p_i}{p_i} -
\frac{1}{2} \left( \frac{q_i - p_i}{p_i} \right)^2 \right ) \\
&= \sum_{i = 1}^{r} q_i \frac{q_i - p_i}{p_i} -
\frac{q_i}{2 p_i} \frac{(q_i - p_i)^2} {p_i}
\end{align*}
Since $\sum_{i = 1}^{r} q_i = \sum_{i = 1}^{r} p_i = 1$,
$\sum_{i = 1}^{r} p_i \frac{q_i - p_i}{p_i} = 0$, so
\begin{align*}
H(Q||P) &\approx \sum_{i=1}^{r} q_i \frac{q_i - p_i}{p_i} - p_i \frac{q_i -
p_i}{p_i} - \frac{q_i}{2 p_i} \frac{(q_i - p_i)^2}{p_i} \\
&= \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}\left(1 - \frac{q_i}{2 p_i}\right) \\
&\approx \frac{1}{2} \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}
\end{align*}
since $q_i \approx p_i$. Multiplying by $n^2 / n^2$ we have,
\begin{align*}
H(Q||P) &\approx \frac{1}{2n} \sum_{i=1}^{r} \frac{(n q_i - n p_i)^2}{n p_i} \\
        &= \frac{1}{2n} \sum_{i=1}^{r} \frac{(X_i - E[X_i])^2}{E[X_i]}
\end{align*}

@@

Stochastic simulations of this, shown in Figure \ref{fig:entropy}, proves when
$P$ is uniform show that the above formula is a very good fit when $n > r$.

\begin{figure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{fig/entropy.pdf}}
\caption{For each value of $n$ plotted, the circles indicate the value of
$H(Q||P)$, averaged over 100 samples of size $n$, where $P$ is uniform and $Q$
is the MLE estimator for $P$ based on $n$ random samples drawn from $r$ bins.
The asterisks and the straight line interpolating them are the theoretical
approximation from Equation \ref{eq:entropy}.
}
\label{fig:entropy}
\end{figure}

@@

The training procedure begins by considering the set of models formed by
including each single nucleotide position from within the training sequences.
For each single nucleotide model, the Bayesian information criterion is
evaluated. If in any of these models the BIC score increases over the empty
model, a non-empty model will be trained, otherwise the training procedure will
halt with an empty model.

@@

Now, the objective function evaluated is the Bayesian information criterion (BIC)
applied to the conditional log-likelihood (CLL).
$$ \text{BIC} = 2 \cdot \text{CLL} - m \log n $$
where CLL in this single nucleotide case can be written as,
$$ \text{CLL} =
\sum_{k \in \{a, c, g, t\}}
\left(
x_k \log \left( \frac{x_k/n}{(x_k + y_k)/n} \right) +
y_k \log \left( \frac{y_k/n}{(x_k + y_k)/n} \right) \right)$$

@@

In summary, if $D_X, D_Y < \frac{3}{2n} \log n$, then $\text{BIC} \le
\text{BIC}_0$ and the empty model is retained. That is, the probability of an
empty model being trained is bounded below by the probability that the KL
divergence of both samples is bounded above by $\frac{3}{2n} \log n$:
$$ \Pr(D_{X} < \frac{3}{2n} \log n) \cdot 
\Pr(D_{Y} < \frac{3}{2n} \log n) $$

@@

In training we consider multiple nucleotides. If $h$ nucleotides are considered,
the probability of a non-empty model being trained is less than
$$ 1 - (\Pr(D < \frac{3}{2n} \log n))^{2h} $$

As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $\chi^2$
distribution with 3 degrees of freedom for large $n$, and so we can estimate
this probability for various values of $n$. This is shown in Figure
\ref{fig:nepr}. From this figure we can see, for example, that training with
more than 10,000 reads results in an incorrect non-empty model with probability
less than 0.0004.

For small $n$, the upper bound on non-empty model probability is much higher.
In our tests, this upper bound is not tight for a small number of reads. We
trained the model on 100 sets of 100 simulated reads, drawn uniformly at random
from exonic sequences and in only 3 trials was a non-empty model trained, far
less than the upper bound expectation of 23 out of 100. This test was repeated
using sets of 10,000 reads, and in no trial was a non-empty model trained.  From
these results, and the results in Supplementary Section 5, we recommend training
our method with at least 10,000 reads.

Here we considered only the case of an empty model versus a non-empty model. A
similar analysis can be conducted to show that if an extraneous position were
added to to the initial empty model, it would be very improbable that more
extraneous positions are edged were added. So that if, despite the very low
probability, a non-empty model were trained from unbiased data, it would be
extremely sparse and have little effect.

\begin{figure}[H]
\centerline{
\includegraphics[width=0.8\textwidth]{fig/nepr.pdf}}
\caption{Plotted is the upper bound on the probability of training a non-empty
model when there is no real bias in the experiment when the model is trained
with various numbers of reads. We fixed the number of nucleotides being
considered as $h = 41$, as was used for the evaluation in Section 3.}
\label{fig:nepr}
\end{figure}
